---
title: "STAT 417 Final Code Work"
author: "Group 8"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

### Initial data pull/clean
```{r}
#read the data from the fed website
library(readr)
url <- "https://fred.stlouisfed.org/graph/fredgraph.csv?id=UNRATENSA"
unrate_data <- read_csv(url)

#convert to time series
unrate_ts <- ts(unrate_data$UNRATENSA, start = c(1948, 1), frequency = 12)
#interpolate the NA values
library(forecast)
unrate_ts_interp <- na.interp(unrate_ts)

#Seasonally adjusted data (for comparison maybe)
urlb <- "https://fred.stlouisfed.org/graph/fredgraph.csv?id=UNRATE"
unrate_data_sa <- read_csv(urlb)

#convert to time series
unratesa_ts <- ts(unrate_data_sa$UNRATE, start = c(1948, 1), frequency = 12)
#interpolate the NA values
library(forecast)
unratesa_ts_interp <- na.interp(unratesa_ts)
```

### ts analysis
```{r}
#full data set and smoothing
plot(unrate_ts_interp, ylab="Unemployment Rate (NSA)", xlab = "time", col = "black", main = "Monthly Unemployment Rate, USA, Not Seasonally Adjusted")
lines(lowess(unrate_ts_interp, f=0.01), lwd=2, col="red")

#stl analysis
unrate_stl <- stl(unrate_ts_interp, s.window = "periodic")
plot(unrate_stl)
season <- unrate_stl$time.series[,1]
trend <- unrate_stl$time.series[,2]
res <- unrate_stl$time.series[,3]

acf(res, main = "ACF(residuals) Unrate data")
```
We can see seasonality in the tiny spikes, probably should blow this graph up on the ppt.

ACF reveals time dependence of the residuals, so we need to model the residuals.

### Residual AR fit model
```{r}
#get the order of the AR fit
fit_ar <- ar.yw(res, order.max = NULL)
fit_ar

#look at the residuals for this new model
fit_ar <- ar.yw(res, order.max = NULL)
fit_resid <- ts(fit_ar$resid)
p<- fit_ar$order
acf(fit_resid[-c(1:p)], main = "ACF AR fit residuals") #remove the first p(order) observations, the rest are res
```

### Trend model
```{r}
time <- 1:length(trend)
dat.trend <- data.frame(trend = trend, time = time)

# trig fit
fit_trig <- lm(trend ~ 
                 cos(2*pi*time/12) + sin(2*pi*time/12) +
                 cos(4*pi*time/12) + sin(4*pi*time/12) +
                 cos(6*pi*time/12) + sin(6*pi*time/12),
               data = dat.trend)
summary(fit_trig)

# polynomial fits
#I() forces literal interpretation of the eqn
fit_poly10 <- lm(trend ~ time + I(time^2) + I(time^3) + I(time^4) + I(time^5) + I(time^6) + I(time^7) + I(time^8) + I(time^9) + I(time^10), data = dat.trend)

fit_poly9 <- lm(trend ~ time + I(time^2) + I(time^3) + I(time^4) + I(time^5) + I(time^6) + I(time^7) + I(time^8) + I(time^9), data = dat.trend)

fit_poly8 <- lm(trend ~ time + I(time^2) + I(time^3) + I(time^4) + I(time^5) + I(time^6) + I(time^7) + I(time^8), data = dat.trend)

summary(fit_poly10)
summary(fit_poly9)
summary(fit_poly8)

# compare trig vs polynomial
#fitted() applies the coefficients to the model for each data point
plot(time, trend, pch=".", main="Trend with Trig and Polynomial fit")
lines(time, fitted(fit_trig), col="red", lwd=2)
lines(time, fitted(fit_poly10), col="blue", lwd=2)
legend("topleft",
       legend=c("Trend", "Trig fit","Poly fit 10"),
       col=c("black", "red", "blue"),
       lwd=2)

# compare polynomial degrees
plot(time, trend, pch=".", main="Trend with Different Polynomial fits")
lines(time, fitted(fit_poly10), col="blue", lwd=2)
lines(time, fitted(fit_poly9), col="orange", lwd=2)
lines(time, fitted(fit_poly8), col="green", lwd=2)
legend("topleft",
       legend=c("Trend", "poly 10","poly 9","poly 8"),
       col=c("black", "blue", "orange","green"),
       lwd=2)

```
Trig fit is not great, the plot is not even close and the p-values are high.
Polynomial fit is good: of the three, degree 10 has the lowest residual standard error. 

### Seasonal model
```{r}
time <- 1:length(season)
dat.season <- data.frame(season = season, time = time)

fit_season1 <- lm(season ~ 
                   cos(2*pi*time/12) + sin(2*pi*time/12),
                 data = dat.season)

summary(fit_season1)

fit_season2 <- lm(season ~ 
                   cos(2*pi*time/12) + sin(2*pi*time/12) +
                   cos(4*pi*time/12) + sin(4*pi*time/12) +
                   cos(6*pi*time/12) + sin(6*pi*time/12),
                 data = dat.season)

summary(fit_season2)

plot(time, season, pch=".", main="Seasonal Component with Trig Fit",
     xlab="Time", ylab="Seasonal Component")
lines(time, fitted(fit_season2), col="red")
legend("topright", legend=c("Seasonal component", "Trig fit"),
       col=c("black", "red"), lwd=2)


```
Graph is super ugly and compressed, but the more complex trig has higher R squared.

### Combined Models
```{r}
#model 1: poly 8 and simple trig
tr1 <- fitted(fit_poly8)

coef_s1 <- coef(fit_season1)

ss1 <- coef_s1[1] +
       coef_s1[2]*cos(2*pi*time/12) +
       coef_s1[3]*sin(2*pi*time/12)

yhat1 <- tr1 + ss1

#model 2: poly 9 and complex trig
tr2 <- fitted(fit_poly9)

coef_s2 <- coef(fit_season2)

ss2 <- coef_s2[1] +
       coef_s2[2]*cos(2*pi*time/12) +
       coef_s2[3]*sin(2*pi*time/12) +
       coef_s2[4]*cos(4*pi*time/12) +
       coef_s2[5]*sin(4*pi*time/12) +
       coef_s2[6]*cos(6*pi*time/12) +
       coef_s2[7]*sin(6*pi*time/12)

yhat2 <- tr2 + ss2

#model 3: ploy 10 and complex trig
tr3 <- fitted(fit_poly10)

yhat3 <- tr3 + ss2

#plot the models
plot(unrate_ts_interp, col="black",
     main="Model Comparison",
     ylab="Unemployment Rate")

lines(ts(yhat1, start=start(unrate_ts_interp), frequency=12),
      col="red")

lines(ts(yhat2, start=start(unrate_ts_interp), frequency=12),
      col="blue")

lines(ts(yhat3, start=start(unrate_ts_interp), frequency=12),
      col="green")

legend("topright",
       legend=c("Actual", "Model1", "Model2", "Model3"),
       col=c("black", "red", "blue", "green"),
       lwd=2)

#only recent years
start_plot <- c(2000,1)

plot(window(unrate_ts_interp, start=start_plot), col="black",
     main="Model Comparison (Recent Period)",
     ylab="Unemployment Rate")

lines(window(ts(yhat1, start=start(unrate_ts_interp), frequency=12), start=start_plot),
      col="red")

lines(window(ts(yhat2, start=start(unrate_ts_interp), frequency=12), start=start_plot),
      col="blue")

lines(window(ts(yhat3, start=start(unrate_ts_interp), frequency=12), start=start_plot),
      col="green")

```


