---
title: "STAT 417 – Smoothing Methods & Forecasting"
author: "Haley Patel (Person A)"
date: "`r Sys.Date()`"
output:
  pdf_document:
    latex_engine: xelatex
    fig_height: 3.5
    fig_width: 7
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.align = "center")
```

---

## Setup: Data & STL Decomposition

```{r data-setup}
library(readr); library(forecast)

# Load UNRATENSA and interpolate NAs
unrate_data      <- read_csv("https://fred.stlouisfed.org/graph/fredgraph.csv?id=UNRATENSA")
unrate_ts        <- ts(unrate_data$UNRATENSA, start = c(1948, 1), frequency = 12)
unrate_ts_interp <- na.interp(unrate_ts)
```

---

## II. Smoothing Methods & Forecasting

### 1. Naïve Baseline

The **seasonal naïve** method repeats the value from the same month one year prior — the simplest
possible benchmark for a seasonal series:

$$\hat{X}_{t+h} = X_{t+h-m}, \quad m = 12$$

```{r naive, fig.cap="Seasonal naïve 24-month forecast (top) and its residuals (bottom)", fig.show='hold', out.width="50%"}
naive_fit   <- snaive(unrate_ts_interp, h = 24)

plot(naive_fit,
     main = "Seasonal Naïve Forecast – UNRATENSA",
     ylab = "Unemployment Rate (NSA)", xlab = "Time")

plot(residuals(naive_fit),
     main = "Residuals – Seasonal Naïve",
     ylab = "Residual", xlab = "Time")
abline(h = 0, col = "red", lty = 2)
```

The residuals show clear structure (large swings around recessions, COVID spike), confirming the naïve
method leaves systematic patterns unexplained. It serves only as a reference benchmark.

---

### 2. Holt-Winters Exponential Smoothing

The Exploratory Data Analysis (EDA) and Seasonal and Trend decomposition using Loess (STL) decomposition established that UNRATENSA has both a time-varying trend and strong
monthly seasonality. This rules out Simple Exponential Smoothing (level only) and Holt's method (trend, no seasonality) as
appropriate models for the raw series. **Holt-Winters (HW)** is therefore the correct exponential
smoothing model, estimating all three components simultaneously:

$$\ell_t = \alpha(X_t - S_{t-m}) + (1-\alpha)(\ell_{t-1} + b_{t-1})$$
$$b_t   = \beta(\ell_t - \ell_{t-1}) + (1-\beta)\,b_{t-1}$$
$$S_t   = \gamma(X_t - \ell_t) + (1-\gamma)\,S_{t-m}$$

where $m = 12$ for monthly data and $\alpha, \beta, \gamma \in (0,1)$ control how quickly each
component adapts to new observations.

```{r hw-fit, fig.cap="Holt-Winters fitted values overlaid on the full UNRATENSA series"}
hw_fit <- HoltWinters(unrate_ts_interp)
hw_fit   # display estimated alpha, beta, gamma

plot(unrate_ts_interp, col = "black",
     main = "Holt-Winters – Fitted vs Observed",
     ylab = "Unemployment Rate (NSA)", xlab = "Time")
lines(hw_fit$fitted[, 1], col = "green", lwd = 1)
legend("topright", legend = c("Observed","HW Fitted"),
       col = c("black","green"), lwd = 2, cex = 0.8)
```

**Result:** $\hat{\alpha} \approx 0.933$, $\hat{\beta} \approx 0.011$, $\hat{\gamma} = 1.000$.
The high $\alpha$ indicates the level estimate tracks recent observations closely; the low $\beta$
means the trend slope changes slowly; $\gamma = 1$ means the seasonal factors are updated
aggressively each month — appropriate for UNRATENSA's stable but precisely-timed seasonal pattern.

---

### 3. Forecast Results

```{r hw-forecast, fig.cap="HW 24-month forecast (full history, top) and zoomed 2015–2028 view (bottom)", fig.show='hold', out.width="50%"}
hw_forecast <- forecast(hw_fit, h = 24)

# Full history view
plot(hw_forecast,
     main = "HW 24-Month Forecast – UNRATENSA",
     ylab = "Unemployment Rate (NSA)", xlab = "Time")

# Zoomed view
recent_start <- c(2015, 1)
plot(window(unrate_ts_interp, start = recent_start),
     xlim = c(2015, end(hw_forecast$mean)[1] + 1),
     ylim = range(c(window(unrate_ts_interp, start = recent_start),
                    hw_forecast$lower, hw_forecast$upper), na.rm = TRUE),
     main = "HW Forecast – 2015 to 2028",
     ylab = "Unemployment Rate (NSA)", xlab = "Time", col = "black")
lines(hw_forecast$mean,        col = "darkgreen", lwd = 2)
lines(hw_forecast$lower[, 2],  col = "darkgreen", lty = 2)
lines(hw_forecast$upper[, 2],  col = "darkgreen", lty = 2)
legend("topright", legend = c("Observed","HW Forecast","95% CI"),
       col = c("black","darkgreen","darkgreen"),
       lty = c(1, 1, 2), lwd = 2, cex = 0.8)
```

The model forecasts unemployment staying near **4.0–4.4%** through early 2028, with the
characteristic seasonal dip in late summer/fall and uptick in winter–spring. Prediction intervals
widen considerably beyond 12 months, reflecting genuine uncertainty over longer horizons.

---

### 4. Accuracy & Residual Diagnostics

```{r accuracy-table}
# Accuracy comparison across methods
cat("=== Seasonal Naïve ===\n");   print(accuracy(naive_fit))
cat("\n=== Holt-Winters ===\n");    print(accuracy(hw_forecast))
```

```{r hw-residuals, fig.cap="HW residual time plot (top) and ACF of residuals (bottom)", fig.show='hold', out.width="50%"}
hw_resid_ts <- ts(na.omit(residuals(hw_forecast)),
                  start = start(unrate_ts_interp), frequency = 12)

plot(hw_resid_ts, main = "HW Residuals",
     ylab = "Residual", xlab = "Time")
abline(h = 0, col = "red", lty = 2)

acf(hw_resid_ts, lag.max = 36, main = "ACF – HW Residuals")
```

```{r ljung-box}
Box.test(hw_resid_ts, lag = 20, type = "Ljung-Box")
```

**Accuracy summary (training set):**

| Method          | RMSE   | MAE    | MAPE   |
|-----------------|--------|--------|--------|
| Seasonal Naïve  | 1.416  | 0.932  | 15.93% |
| Holt-Winters    | **0.454**  | **0.226**  | **4.13%** |

Holt-Winters achieves substantially lower error on all metrics. The residual ACF shows remaining autocorrelation
(Ljung-Box p < 0.001), indicating the residuals are not white noise — this motivates the subsequent
ARIMA modeling step to capture the remaining dependence structure.

---

### Summary

Holt-Winters outperforms the naïve baseline by explicitly modeling UNRATENSA's
trend and seasonality simultaneously. The estimated parameters ($\alpha \approx 0.93$,
$\beta \approx 0.01$, $\gamma = 1$) are economically interpretable: the unemployment level
shifts quickly, the trend evolves slowly, and the seasonal pattern is recalibrated every month.
The residual diagnostics confirm that HW provides a meaningful improvement over naïve methods,
while the remaining autocorrelation in residuals sets up the ARIMA analysis (Connor) to build on
this foundation.
